Automatic prostate cancer detection through \acs*{dce}-\acs*{mri} images: all you need is a good normalization

### 2016-08-27 01_introduction/introduction.tex
finished

In this work, we propose a fully automatic normalization method for \ac{dce}-\ac{mri} that reduces the inter-patient variability of the data.
{\color{red}The benefit of our approach will be shown while using quantitative and semi-quantitative approaches.
Additionally, we show that using the whole normalized \ac{dce}-\ac{mri} signal is preferable to quantitative and semi-quantitative methods, leading to the best classification performance.}

>>>>

In this work, we propose a fully automatic normalization method for \ac{dce}-\ac{mri} that reduces the inter-patient variability of the data.
The benefit and simplicity of our approach will be shown by classifying the whole normalized \ac{dce}-\ac{mri} signal and comparing with the state-of-the-art quantitative and semi-quantitative methods.
Additionally, we will show that using this normalization approach in conjunction with the quantitative methods improves the classification performance in for most of the models.

{\color{red}The benefit of our approach will be shown while using quantitative and semi-quantitative approaches.
Additionally, we show that using the whole normalized \ac{dce}-\ac{mri} signal is preferable to quantitative and semi-quantitative methods, leading to the best classification performance.}

### 2016-08-27 02_methods/methods.tex
finished

In \ac{dce}-\ac{mri}, the intensity \ac{pdf} of prostate gland does not follow a unique type of distribution such as Rician or Gaussian distribution, as shown in Fig.\,\ref{subfig:pathist}.
Indeed, the inter-patient variations are more complex due to the temporal acquisition.
A better representation to observe these variations is to represent the intensity \ac{pdf} of the \ac{dce}-\ac{mri} data over time using a heatmap representation as shown in Fig.\,\ref{subfig:pathhist}.
Analyzing this heatmap representation across patients (see Fig.\,\ref{subfig:pat2}), the following variations are highlighted:
(i) an intensity offset $\Delta_i$ of the \ac{pdf} peak at pre-contrast,
(ii) a time offset $\Delta_t$ depending of the contrast agent arrival, and
(iii) a change of scale $\sigma_i$ related to the signal enhancement.
Therefore, our normalization method should attenuate all these variations and be performed globally across the different time sequence rather than for each independent sequence.

\subsubsection{Graph-based intensity offset correction}

@@

Before to standardize each sequence, the first step of the normalization is to cancel the intensity specific at each patient, occurring due to the media injection.
As previously mentioned, the intensity \ac{pdf} does not always follow either a Rician or a Gaussian distribution over time, in \ac{dce}-\ac{mri}.
Therefore, the mean of these distributions cannot be used as a potential estimate for these offsets.
Additionally, these offsets should be characterized by a smooth transition between series over time.
Thus, this problem is solved using the graph-theory: considering the intensity \ac{pdf} over time as shown in Fig.\,\ref{subfig:heatmaprep}, the offsets correspond to the boundary splitting the heatmap in two partitions such that they are as close as possible to the peak of the intensity \ac{pdf} (see Fig.\,\ref{fig:estimator} for an illustration).
Given the heatmap, a directed weighted graph $\mathcal{G}=(\mathcal{V}, \mathcal{E})$ is built by taking each bar--- i.e., the probability for a given time and pixel intensity---of the heatmap as a node and connecting each pair of bars by an edge.
The edge weight $w_{ij}$ between two nodes $i$ and $j$ corresponding to two pixels at position $(x_i, y_i)$ and $(x_j, y_j)$, respectively, is defined as in Eq.\,\eqref{eq:weight}:

@@

Illustration of the correction applied to each \ac{rmse} patient is shown in Fig.\,\ref{fig:rmseal}.
Once all these parameters have been inferred, the data can be shifted as well as scale.

@@

As stated in the introduction, one of our main challenge is to demonstrate the benefit of using our normalization method prior to extract parameters using quantitative and semi-quantitative methods.
In this section, we summarize the different methods which have been used for the quantification of \ac{dce}-\ac{mri} for \ac{cap} detection~\citep{lemaitre2015computer}.
Furthermore, we would like to emphasize the following additional contributions: (i) a novel automatic \ac{aif} estimation algorithm based on clustering and (ii) a simplified semi-quantitative method using constrained optimization.

>>>>

In \ac{dce}-\ac{mri}, the intensity \ac{pdf} of prostate gland does not follow a unique type of distribution such as Rician or Gaussian distribution, as shown in Fig.\,\ref{subfig:pathhist}.
Indeed, the inter-patient variations are more complex due to the temporal acquisition.
A better representation to observe these variations is to represent the intensity \ac{pdf} of the \ac{dce}-\ac{mri} data over time using a heatmap representation as shown in Fig.\,\ref{subfig:pathhist}.
Analyzing this heatmap representation across patients (see Fig.\,\ref{subfig:pat2}), the following variations are highlighted:
(i) an intensity offset $\Delta_i$ of the \ac{pdf} peak at pre-contrast,
(ii) a time offset $\Delta_t$ depending of the contrast agent arrival, and
(iii) a change of scale $\sigma_i$ related to the signal enhancement.
Therefore, our normalization method should attenuate all these variations and be performed globally across the different time sequence rather than for each independent sequence.

\subsubsection{Graph-based intensity offsets correction}\label{sec:intoffsets}

@@

Before to standardize each sequence, the first step of the normalization is to cancel the intensity specific at each patient, occurring due to the media injection.
As previously mentioned, the intensity \ac{pdf} does not always follow either a Rician or a Gaussian distribution over time, in \ac{dce}-\ac{mri}.
Therefore, the mean of these distributions cannot be used as a potential estimate for these offsets.
Additionally, these offsets should be characterized by a smooth transition between series over time.
Thus, this problem is solved using the graph-theory: considering the intensity \ac{pdf} over time as shown in Fig.\,\ref{subfig:pathhist}, the offsets correspond to the boundary splitting the heatmap in two partitions such that they are as close as possible to the peak of the intensity \ac{pdf} (see Fig.\,\ref{fig:estimator} for an illustration).
Given the heatmap, a directed weighted graph $\mathcal{G}=(\mathcal{V}, \mathcal{E})$ is built by taking each bar--- i.e., the probability for a given time and pixel intensity---of the heatmap as a node and connecting each pair of bars by an edge.
The edge weight $w_{ij}$ between two nodes $i$ and $j$ corresponding to two pixels at position $(x_i, y_i)$ and $(x_j, y_j)$, respectively, is defined as in Eq.\,\eqref{eq:weight}:

@@

Illustration of the correction applied to each \ac{rmse} patient is shown in Fig.\,\ref{fig:rmseal}.
Once all these parameters have been inferred, the data are shifted as well as scale.

The resulting normalized data can be classified into two fashions: (i) each normalized signal can be classified considering the whole \ac{dce}-\ac{mri} signal or (ii) the normalized data can be fitted using a quantitative method, as presented in the next section.
However, in the second strategy, this is necessary to apply common intensity offsets such that the data follow a shape as expected by the different quantitative models.
The set of offsets applied is in fact corresponding to the maximum offsets found in Sect.\,\ref{sec:intoffsets}.

@@

In this section, we summarize the different methods which have been used for the quantification of \ac{dce}-\ac{mri} for \ac{cap} detection~\citep{lemaitre2015computer} and which will be used for comparison in this work.
Furthermore, we would like to emphasize the following additional contributions: (i) a novel automatic \ac{aif} estimation algorithm based on clustering and (ii) a simplified semi-quantitative method using constrained optimization.

### 2016-08-27 03_experiments/experiments.tex
finished

\subsection{Classification of individual parameter for each model}

% \begin{table}
%   \caption{\acs*{auc} of the individual features for each method.}
%   \centering
%   \begin{tabular}{l c c c c}
%     \toprule
%     \textbf{Features} & \multicolumn{2}{c}{Un-normalized data} & \multicolumn{2}{c}{Normalized data} \\
%     & \acs*{rf} & \acs*{nb} & \acs*{rf} & \acs*{nb} \\
%     \midrule
%     \textbf{Brix model} & & & & \\
%     \quad $A$         & 0.54 & 0.62 & 0.58 & 0.67 \\
%     \quad $k_{el}$    & 0.55 & 0.52 & 0.54 & 0.61 \\
%     \quad $k_{ep}$    & 0.51 & 0.52 & 0.51 & 0.58 \\
%     \textbf{Hoffmann model} & & & & \\
%     \quad $A$         & 0.52 & 0.50 & 0.51 & 0.56 \\
%     \quad $k_{el}$    & 0.55 & 0.53 & 0.54 & 0.64 \\
%     \quad $k_{ep}$    & 0.55 & 0.50 & 0.53 & 0.66 \\
%     \textbf{Tofts model with population \ac{aif}} & & & & \\
%     \quad $K_{trans}$ & 0.56 & 0.62 & 0.56 & 0.65 \\
%     \quad $v_{e}$     & 0.51 & 0.50 & 0.50 & 0.52 \\
%     \quad $v_{p}$     & 0.53 & 0.63 & 0.55 & 0.53 \\
%     \textbf{Tofts model with patient \ac{aif}} & & & & \\
%     \quad $K_{trans}$ & 0.57 & 0.66 & 0.56 & 0.65 \\
%     \quad $v_{e}$     & 0.49 & 0.50 & 0.51 & 0.52 \\
%     \quad $v_{p}$     & 0.53 & 0.37 & 0.57 & 0.65 \\
%     \textbf{\ac{pun} model} & & & & \\
%     \quad $a_0$       & 0.52 & 0.53 & 0.53 & 0.51  \\
%     \quad $r$         & 0.53 & 0.59 & 0.55 & 0.55 \\
%     \quad $\beta$     & 0.55 & 0.56 & 0.53 & 0.44 \\
%     \textbf{Semi-quantitative analysis} & & & & \\
%     \quad wash-in     & 0.59 & 0.64 & 0.55 & 0.51 \\
%     \quad wash-out    & 0.52 & 0.50 & 0.56 & 0.66 \\
%     \quad IAUC        & 0.51 & 0.61 & 0.52 & 0.64 \\
%     \quad $\tau$      & 0.57 & 0.57 & 0.56 & 0.61 \\
%     \quad $S_M - S_0$ & 0.56 & 0.63 & 0.53 & 0.64 \\
%     \bottomrule
%   \end{tabular}
%   \label{tab:resfeats}
% \end{table}

\begin{table}
  \caption{\acs*{auc} for each individual pharmacokinetic parameter using a \acs*{nb} classifier.}
  \centering
  \begin{tabular}{l c c}
    \toprule
    \textbf{Features} & Un-normalized data & Normalized data \\
    \midrule
    \textbf{Brix model} & & \\
    \quad $A$         & 0.62 & 0.67 \\
    \quad $k_{el}$    & 0.52 & 0.61 \\
    \quad $k_{ep}$    & 0.52 & 0.58 \\
    \textbf{Hoffmann model} & & \\
    \quad $A$         & 0.50 & 0.56 \\
    \quad $k_{el}$    & 0.53 & 0.64 \\
    \quad $k_{ep}$    & 0.50 & 0.66 \\
    \textbf{Tofts model with population \ac{aif}} & & \\
    \quad $K_{trans}$ & 0.62 & 0.65 \\
    \quad $v_{e}$     & 0.50 & 0.52 \\
    \quad $v_{p}$     & 0.63 & 0.53 \\
    \textbf{Tofts model with patient \ac{aif}} & & \\
    \quad $K_{trans}$ & 0.66 & 0.65 \\
    \quad $v_{e}$     & 0.50 & 0.52 \\
    \quad $v_{p}$     & 0.37 & 0.65 \\
    \textbf{\ac{pun} model} & & \\
    \quad $a_0$       & 0.53 & 0.51 \\
    \quad $r$         & 0.59 & 0.55 \\
    \quad $\beta$     & 0.56 & 0.44 \\
    \textbf{Semi-quantitative analysis} & & \\
    \quad wash-in     & 0.64 & 0.51 \\
    \quad wash-out    & 0.50 & 0.66 \\
    \quad IAUC        & 0.61 & 0.64 \\
    \quad $\tau$      & 0.57 & 0.61 \\
    \quad $S_M - S_0$ & 0.63 & 0.64 \\
    \bottomrule
  \end{tabular}
  \label{tab:resfeats}
\end{table}

This experiment consists in assessing the benefit of normalizing the \ac{dce}-\ac{mri}, prior to classify each individual parameter for each model.
Therefore, a Gaussian \ac{nb} classifier is used in conjunction with a \ac{lopo}.
The results are summarized in Table~\ref{tab:resfeats}.
In general, data normalization leads to better classification performance while extracting pharmacokinetic related parameters.
Only the \ac{pun} model does not follow this tendency.
{\color{red} Find out why!!!}

\subsection{Classification by combining the parameters for each model}

\begin{figure}
  \centering
  \hspace*{\fill}
  \subfigure[Without normalization.]{\label{fig:rfpharmaunorm}\includegraphics[width=.49\textwidth]{03_experiments/figures/unormalized/rf.pdf}} \hfill
  \subfigure[With normalization.]{\label{fig:rfpharmanorm}\includegraphics[width=.49\textwidth]{03_experiments/figures/normalized/rf.pdf}}
  \hspace*{\fill}
  \caption{\acs*{roc} analysis using a \acs*{rf} classifier with and without normalization \ac{dce}-\ac{mri} data for different pharmacokinetic models.}
  \label{fig:normpharmarf}
\end{figure}

In previous proposed \ac{cad} systems, the pharmacokinetic of a single model are combined.
Therefore, for each model, all parameters are combined and classified using a \ac{rf} classifier in a \ac{lopo} fashion.
The use of \ac{rf} is motivated since that it leads to the best performance in the state-of-the-art methods~\citep{litjens2014computer}.
Results are summarized by performing a \ac{roc} analysis and computing the \ac{auc}, as reported in Fig.\,\ref{fig:normpharmarf}.
Independently to the quantification model, the data normalization improves the\ac{auc} and classification performance.

% \begin{figure}
%   \centering
%   \hspace*{\fill}
%   \subfigure[Without normalization.]{\label{fig:nbpharmaunorm}\includegraphics[width=.49\textwidth]{03_experiments/figures/unormalized/nb.pdf}} \hfill
%   \subfigure[With normalization.]{\label{fig:nbpharmanorm}\includegraphics[width=.49\textwidth]{03_experiments/figures/normalized/nb.pdf}}
%   \hspace*{\fill}
%   \caption{\acs*{roc} analysis using a \acs*{nb} classifier with and without normalization \ac{dce}-\ac{mri} data for different pharmacokinetic models.}
%   \label{fig:normpharmanb}
% \end{figure}

% \begin{figure}
%   \centering
%   \hspace*{\fill}
%   \subfigure[\acs*{rf} classifier.]{\label{fig:rfunorm}\includegraphics[width=.49\textwidth]{03_experiments/figures/rf.pdf}} \hfill
%   \subfigure[\acs*{nb} classifier.]{\label{fig:nbunorm}\includegraphics[width=.49\textwidth]{03_experiments/figures/nb.pdf}}
%   \hspace*{\fill}
%   \caption{\acs*{roc} analysis using the entire \ac{dce}-\ac{mri} signal with and without normalization.}
%   \label{fig:unormdcesignal}
% \end{figure}

@@

As stated in the introduction, the quantification methods are extracting a set of parameters characterizing the enhancement \ac{dce}-\ac{mri} signal.
However, this extraction could lead to a loss of information.
This experiment is performed to assess if making use of the whole \ac{dce}-\ac{mri} signal instead of the just the pharmacokinetic parameters can improve the classification performance.
Therefore, each enhanced \ac{dce}-\ac{mri} signal, normalized and un-normalized, is classified using a \ac{rf} classifier in a \ac{lopo} fashion.
The \ac{roc} analysis and \ac{auc} are reported in Fig.\,\ref{fig:rfnormdcesignal}.
It can be noted that the worst performance are achieved while classifying raw data without normalization.
However, data normalization in conjunction with the use of the whole \ac{dce}-\ac{mri} signal is the strategy which outperforms all others.

>>>>

\subsection{Goodness of model fitting}\label{sec:fit}

%{\color{red} In case that we have issue with $R^2$, we need to provide the AIC since that the model are usually non-linear.}

\begin{table*}
  \caption{Coefficient of determination $R^{2}$ (i.e., $\mu \ (\pm \sigma)$), while fitting data with the different quantification models.}
  \centering
  \resizebox{\columnwidth}{!}{
  \begin{tabular}{lc c c c c c}
    \toprule
    Data type & Brix & Hoffmann & Tofts population \ac{aif} & Tofts patient \ac{aif} & \ac{pun} & Semi-quantitative \\
    \midrule
    Un-normalized & $0.85 \ (\pm 0.11)$ & $0.81 \ (\pm 0.17)$ & $0.84 \ (\pm 0.14)$ & $0.88 \ (\pm 0.12)$ & $0.27 \ (\pm 0.18)$ & $0.64 \ (\pm 0.24)$  \\
    Normalized    & $0.92 \ (\pm 0.05)$ & $0.72 \ (\pm 0.32)$ & $0.92 \ (\pm 0.06)$ & $0.90 \ (\pm 0.10)$ & $0.28 \ (\pm 0.20)$ & $0.75 \ (\pm 0.20)$  \\
    \bottomrule
  \end{tabular}
  }
  \label{tab:r2}
\end{table*}

Parameter estimation of the quantification methods are related to fit a specific model to the \ac{dce}-\ac{mri} data.
Therefore, this section report the goodness of fitting by computing the coefficient of determination $R^2$ such as in Eq.\,\eqref{eq:r2}

\begin{equation}
  R^2 = 1 - \frac{\sum_{t = 0}^{T} (s_t - \hat{s}_t)^2}{\sum_{t = 0}^{T} (s_t - \bar{s})^2} ,
  \label{eq:r2}
\end{equation}

\noindent where $s_t$ and $\hat{s}_t$ are the signal to be fitted and the estimated signal at time $t$, respectively; $\bar{s}$ is the average signal to be fitted.

Mean and standard-deviation of the coefficient of determination $R^{2}$ is reported in Table~\ref{tab:r2} for each quantification model.
Brix, Hoffmann, and Tofts models are fitted with a coefficient $R^{2}$ superior to 0.80.%, implying that they are adequate models for that purpose.
Additionally, the proposed \ac{pun} model does not seem to fit well the data.
Normalizing the data improve the coefficient $R^2$ for all the methods apart of the Hoffmann model.
The large standard deviation for this model might imply that there is some cases, that the fitting fails.

\subsection{Detection of \acs*{cap} using pharmacokinetic parameters}

\begin{table*}
  \caption{\acs*{auc} for each individual pharmacokinetic parameter using a \acs*{rf} classifier.}
  \centering
  \resizebox{\columnwidth}{!}{
  \begin{tabular}{l c c}
    \toprule
    \textbf{Features} & Un-normalized data & Normalized data \\
    \midrule
    \textbf{Brix model} & & \\
    \quad $A$         & 0.54 & 0.55 \\
    \quad $k_{el}$    & 0.51 & 0.50 \\
    \quad $k_{ep}$    & 0.55 & 0.58 \\
    \textbf{Hoffmann model} & & \\
    \quad $A$         & 0.52 & 0.51 \\
    \quad $k_{el}$    & 0.55 & 0.53 \\
    \quad $k_{ep}$    & 0.55 & 0.54 \\
    \textbf{Tofts model with population \ac{aif}} & & \\
    \quad $K_{trans}$ & 0.56 & 0.55 \\
    \quad $v_{e}$     & 0.49 & 0.49 \\
    \quad $v_{p}$     & 0.53 & 0.50 \\
    \textbf{Tofts model with patient \ac{aif}} & & \\
    \quad $K_{trans}$ & 0.57 & 0.57 \\
    \quad $v_{e}$     & 0.51 & 0.53 \\
    \quad $v_{p}$     & 0.53 & 0.55 \\
    \textbf{\ac{pun} model} & & \\
    \quad $a_0$       & 0.52 & 0.53 \\
    \quad $r$         & 0.55 & 0.57 \\
    \quad $\beta$     & 0.53 & 0.55 \\
    \textbf{Semi-quantitative analysis} & & \\
    \quad wash-in     & 0.57 & 0.53 \\
    \quad wash-out    & 0.52 & 0.49 \\
    \quad IAUC        & 0.51 & 0.51 \\
    \quad $\tau$      & 0.57 & 0.54 \\
    \quad $S_M - S_0$ & 0.56 & 0.53 \\
    \bottomrule
  \end{tabular}
  }
  \label{tab:resfeats}
\end{table*}

\begin{figure*}
  \centering
  \hspace*{\fill}
  \subfigure[Without normalization.]{\label{fig:rfpharmaunorm}\includegraphics[width=.49\textwidth]{03_experiments/figures/unormalized/rf.pdf}} \hfill
  \subfigure[With normalization.]{\label{fig:rfpharmanorm}\includegraphics[width=.49\textwidth]{03_experiments/figures/normalized/rf.pdf}}
  \hspace*{\fill}
  \caption{\acs*{roc} analysis using a \acs*{rf} classifier with and without normalization \ac{dce}-\ac{mri} data for different pharmacokinetic models.}
  \label{fig:normpharmarf}
\end{figure*}

To study the potential benefit of our normalization, \ac{cap} are detected using pharmacokinetic parameters estimated from un-normalized and normalized \ac{dce}-\ac{mri} data.
Each individual pharmacokinetic parameter is classified to evaluate their individual discriminative power to detect \ac{cap}.
Therefore, a \ac{rf} classifier is used in conjunction with a \ac{lopo}.
The use of \ac{rf} is motivated since that it leads to the best performance in the state-of-the-art methods~\citep{litjens2014computer,lemaitre2015computer}.
Results are summarized in Table~\ref{tab:resfeats} in terms of \ac{auc}.
Normalization can improve the detection of \ac{cap}; however, the benefit of normalization is more obvious by combining the pharmacokinetic features of each model, as previously done in traditional \ac{cad} system~\citep{lemaitre2015computer}.
For the latter configuration, results are summarized by performing a \ac{roc} analysis and computing the \ac{auc}, as reported in Fig.\,\ref{fig:normpharmarf}.
Quantification using normalized data outperforms quantification using un-normalized data in terms of classification performance apart of Hoffmann and Tofts population-based \ac{aif} models.
The reasons behind the decrease of the \ac{auc} might be related to: (i) a poor fitting as discussed in Sect.\,\ref{sec:fit} (cf., Hoffmann model) and (ii) a small number of patient while estimating some parameters (cf., Tofts model).
The best classification performance are obtained using the semi-quantitative approach with an \ac{auc} of 0.655.

@@

As stated in the introduction, the quantification methods are extracting a set of parameters characterizing the enhancement \ac{dce}-\ac{mri} signal.
However, this extraction might lead to a loss of information.
This experiment is performed to assess if making use of the whole \ac{dce}-\ac{mri} signal instead of the just the pharmacokinetic parameters can improve the classification performance.
Therefore, each enhanced \ac{dce}-\ac{mri} signal, normalized and un-normalized, is classified using a \ac{rf} classifier in a \ac{lopo} fashion.
The \ac{roc} analysis and \ac{auc} are reported in Fig.\,\ref{fig:rfnormdcesignal}.
Classification without normalization lead to the worst performance, with an \ac{auc} of 0.568.
However, data normalization in conjunction with the use of the whole \ac{dce}-\ac{mri} signal is the strategy which outperforms all others, with an \ac{auc} of 0.666.
